---
title: "677finalproject"
author: "Bella"
date: "2024-05-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Empirical Bayes Methods in Modern Statistics

## 1. Main Points and Computational Methods

Empirical Bayes methods provide a robust statistical framework that merges the strengths of traditional Bayesian inference with empirical data-driven approaches. This chapter delves into the computational advancements that have transformed statistical practice by integrating modern computational power with Bayesian principles.

### The Shift from Classical to Computational Statistics

After World War II, the introduction of electronic computation revolutionized data analysis, enabling the development of more dynamic statistical models that could handle larger datasets and more complex analyses. Classical statistics, heavily reliant on assumptions of normal distributions and simple random samples, began to give way to methods that could utilize the vast amounts of data being generated in the modern era.



### Robbins' Formula: A Keystone of Empirical Bayes

Herbert Robbins introduced a formula that epitomizes the empirical Bayes approach. This formula allows for the estimation of parameters in environments where the prior distribution is not known but can be estimated from the data itself. Robbins’ approach was particularly innovative in that it provided a method to estimate the number of unseen events—a problem classically considered in ecological and biological studies for estimating species richness.



#### Computational Demonstration of Robbins' Formula

Consider the problem of estimating the rate of insurance claims based on historical data. Robbins' formula uses the empirical data to update prior beliefs about the claims rate, thereby refining predictions for future claims. This method has profound implications in industries like insurance and healthcare, where predicting future events accurately is crucial for resource allocation and planning.



### Practical Application Scenarios

- **Insurance Claims Prediction:** Using past claim data to forecast future claims.
- **Medical Studies:** Estimating disease incidence from observed data in clinical trials.



## 2. Mathematics Underlying the Material

This section explores the rigorous mathematical frameworks that underpin Empirical Bayes methods, highlighting how these methods integrate empirical data with Bayesian statistical principles to provide more accurate and reliable estimates.

### Bayesian Statistics: The Foundation

At its core, Bayesian statistics involves updating the probability for a hypothesis as more evidence or information becomes available. This process incorporates prior knowledge (prior distribution) and new data (likelihood) to produce a revised probability (posterior distribution).

#### Key Concepts in Bayesian Analysis

- **Prior Distribution:** Represents what is known about the parameter before considering the current data.
- **Likelihood:** The probability of the data given the parameters, which helps in updating the beliefs based on new evidence.
- **Posterior Distribution:** The result of the Bayesian update process, combining the prior and the likelihood.

### Deep Dive into the Poisson-Gamma Model

A frequently used model in Empirical Bayes settings is the Poisson-Gamma model, where the prior distribution for a Poisson rate parameter is assumed to be Gamma. This model is particularly useful in scenarios where events happen independently at a constant but unknown rate, such as the number of claims or the occurrence of a particular disease.

#### Mathematical Derivation

Let \(\lambda\) be the rate parameter of a Poisson distribution modeling the number of events. Assuming a Gamma distribution as the prior for \(\lambda\) with shape \(\alpha\) and rate \(\beta\), the posterior distribution after observing data also follows a Gamma distribution but with updated parameters.

```{r echo=FALSE}
# Gamma-Poisson Update Example in R
alpha_prior <- 2    # shape parameter of the prior
beta_prior <- 1     # rate parameter of the prior
data <- c(3, 2, 5)  # observed counts

alpha_post <- alpha_prior + sum(data)
beta_post <- beta_prior + length(data)
```

# Posterior mean
posterior_mean <- alpha_post / beta_post


## 3. Mathematical Underpinnings

This section explores the mathematical foundations of Empirical Bayes methods, emphasizing the integration of Bayesian statistics with empirical data. Key concepts include the Bayesian updating mechanism, where prior distributions are adjusted in light of new, empirical data, enhancing the accuracy of statistical estimates.

### Bayesian Analysis

Bayesian analysis combines prior knowledge with observed data through a rigorous mathematical framework, providing a comprehensive way to update beliefs based on evidence. This approach is central to Empirical Bayes methods, which use data to refine and update prior distributions.

## 4. Historical Context and Evolution

The chapter provides a historical overview of the development of Bayesian and Empirical Bayes methodologies, noting significant contributions by statisticians such as Robbins. These developments underscored a broader trend towards more data-centric statistical practices, which were made possible by advancements in computational technology.

### Contributions of Key Figures

Significant figures in the field of Bayesian statistics, such as Robbins, have contributed foundational concepts and methodologies that have shaped the practice of modern statistics. Their work has facilitated the broader acceptance and application of Bayesian methods across various scientific domains.

## 5. Implications for Statistical Practice

Empirical Bayes methods are not just academically interesting; they have profound practical implications. They are particularly useful in the context of big data and complex models where traditional statistical methods might be insufficient.

### Modern Applications1

Today, Empirical Bayes methods are applied in a wide array of fields from genomics to machine learning, where they help in handling large datasets and complex predictive models, offering a robust framework for making informed decisions based on incomplete data.


# R Script for Demonstration



```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary library
library(tidyverse)

# Simulate some data: Let's say we have historical data of insurance claims for 1000 policyholders
set.seed(123)
claims_data <- rpois(1000, lambda = 2)  # Poisson distribution, average 2 claims per policyholder

# Calculate the empirical prior based on the observed data
empirical_prior <- density(claims_data)

# Plot the empirical prior
plot(empirical_prior, main = "Empirical Prior Density", xlab = "Number of Claims", ylab = "Density")

# Assume new data comes in
new_claims_data <- rpois(1000, lambda = 2.5)  # Slightly higher average due to some external factor



```


_data) * length(claims_data) + sum(new_claims_data)) / (length(claims_data) + length(new_claims_data))

###New Prior distribution visualization
new_empirical_prior <- density(c(claims_data, new_claims_data)) # Combine old and new data for updating
plot(new_empirical_prior, main = "Updated Empirical Prior Density", xlab = "Number of Claims", ylab = "Density")

###Example of making predictions
Suppose we want to predict the number of claims for a new set of 1000 policyholders
predicted_claims <- rpois(1000, lambda = updated_lambda)
summary(predicted_claims) # Basic summary statistics of the predicted claims

###Visualize the predictions
hist(predicted_claims, breaks = 30, main = "Predicted Number of Claims", xlab = "Number of Claims", col = "skyblue")




### Explanation of the R Script:
- **Data Simulation:** This script starts by simulating historical data for insurance claims using a Poisson distribution, reflecting typical claim frequencies.
- **Empirical Prior:** An empirical prior density is estimated from the historical data, visualized to provide an intuitive understanding of the claim distribution among policyholders.
- **Bayesian Updating:** New data simulating a slight increase in claims is then used to update the prior distribution, demonstrating how Empirical Bayes adjusts to new information.
- **Prediction:** Finally, predictions are made for future claims based on the updated model, and the distribution of these predictions is visualized.

### Using the R Markdown and Script:
To use the provided R Markdown and script:
1. **Set Up:** Ensure you have R and RStudio installed, and the necessary packages (like `tidyverse`) are available.
2. **Run the Markdown File:** Copy the markdown content into an R Markdown file (`.Rmd`), and knit the document to generate a nicely formatted report in HTML or PDF format.
3. **Execute the Script:** Run the R script in RStudio to see how empirical Bayesian methods can be practically applied and visualized.

This structured approach not only elucidates the theoretical aspects of Empirical Bayes but also demonstrates its practical application, making your project comprehensive and informative.



### Modern Applications2

# Statistical Practice Implications

Empirical Bayes methods, particularly Robbins' Formula, have wide-ranging applications across various domains, offering powerful solutions for modern data-driven challenges. These methods help in refining predictions, optimizing decision-making processes, and managing uncertainties in estimates, which are crucial for effective planning and resource management in multiple fields.

## Applications in Real-World Statistical Practices

Empirical Bayes methods have been pivotal in shaping decision-making processes where prior data is available but incomplete, and where future predictions need to be made with a degree of confidence. Robbins’ Formula, in particular, serves as a robust tool for updating estimates based on new evidence.

### Example: Resource Allocation in Healthcare

Consider the management of hospital resources, which is critical to healthcare planning. Robbins' Formula can be used to predict hospital readmissions, which are a key factor in determining the required number of beds, staff levels, and medical supplies.

#### Scenario: Predicting Hospital Readmissions

Suppose a hospital has historical data on patient readmissions for various diseases. By using Robbins' Formula, the hospital can estimate the probability of readmissions for the upcoming year based on past data, adjusting for any changes in hospital policy, demographics, or treatment methods.

```{r echo=FALSE}
# Sample R code to demonstrate the application of Robbins' Formula in predicting hospital readmissions

# Simulate some historical readmission data (Poisson distributed)
set.seed(202)
historical_readmissions <- rpois(500, lambda = 1.5)  # Historical average of 1.5 readmissions per patient

# Assume the prior distribution of the lambda parameter is Gamma
alpha_prior <- 10
beta_prior <- 7

# Update the prior based on historical data
alpha_post <- alpha_prior + sum(historical_readmissions)
beta_post <- beta_prior + length(historical_readmissions)

# Calculate the posterior mean (estimate for next year's readmission rate)
lambda_estimate <- alpha_post / beta_post

# Use this estimate to plan resources
expected_readmissions <- lambda_estimate * 600  # Predicting for 600 patients

print(paste("Expected number of readmissions for next year:", round(expected_readmissions, 0)))
```


Educational Testing Services
Another compelling application of Robbins' Formula is in educational settings, such as predicting student performance on standardized tests based on previous years' data. Educational administrators can use these predictions to tailor instructional methods, allocate educational resources more effectively, and identify areas where intervention is necessary.

Implementation in an Educational Context
Using past performance data, schools can estimate the variance in test scores and adjust teaching resources to target areas where students historically struggle. This predictive approach ensures that resources are directed where they are most needed, potentially improving overall student performance and educational outcomes.

Conclusion
These examples illustrate the practical utility of Empirical Bayes methods in real-world applications. By incorporating empirical data into Bayesian frameworks, Robbins’ Formula and similar techniques allow organizations to make better-informed decisions, leveraging historical data to anticipate future needs and challenges.



